Final paper: Towards Proving Termination of \texttt{zelda-mosaic} Algorithms under Conditions on the Input

### 2021-05-02 paper/paper.tex
paper: finish the theorems

\paragraph{Linearizing a matrix} See \S~\ref{S:linearized_indexing} for more
details on matrix linearization. To define linearization in Coq, we use a
two-stage unpacking as with other matrix operations. The function
\(\func{linearize'}\) turns scalars into a singleton list and recursively
flat-maps itself over matrices to produce the one-dimensional matrix as a list.
The function \(\func{linearize}\) unpacks a matrix, then transforms the contents
by wrapping the result of mapping \(\func{linearize'}\) into \(\func{Scalar}\)s
in a \(\func{Matrix}\) and transforms the shape by taking the product of the
(normalized) shape of the matrix. Recall that the length of the linearized
vector is the product of the sizes of all the dimensions. The definitions are
spelled out below:

@@

\begin{lem}\label{Lem:wfI_all_wf_t}
    If all elements of the list \(\var{ms}\) are well-formed (using the inductive
    definition) with respect to some shape \(t\), then the following matrix is
    also well-formed: \(\{ \var{shape} = |\var{ms}| :: t; \var{contents} =
    \func{Matrix}\, \var{ms} \}\).
\end{lem}

@@

\begin{thm}[Linearize has resulting shape the product of the original]\label{Th:linearize_product}
\end{thm}

\begin{thm}[Linearize preserves the well-formed property]\label{Th:linearize_wf}
\end{thm}

>>>>

\paragraph{Linearizing a matrix}\label{S:linearize_coq} See
\S~\ref{S:linearized_indexing} for more details on matrix linearization. To
define linearization in Coq, we use a two-stage unpacking as with other matrix
operations. The function \(\func{linearize'}\) turns scalars into a singleton
list and recursively flat-maps itself over matrices to produce the
one-dimensional matrix as a list. The function \(\func{linearize}\) unpacks a
matrix, then transforms the contents by wrapping the result of mapping
\(\func{linearize'}\) into \(\func{Scalar}\)s in a \(\func{Matrix}\) and
transforms the shape by taking the product of the (normalized) shape of the
matrix. Recall that the length of the linearized vector is the product of the
sizes of all the dimensions. The definitions are spelled out below:

@@

\begin{lem}\label{Lem:wfI_all_wf_t}
    If all elements of the list \(\var{ms}\) are well-formed (using the inductive
    definition) with respect to some shape \(t\), then the following matrix is
    also well-formed in the inductive definition: \(\{ \var{shape} = |\var{ms}|
    :: t; \var{contents} = \func{Matrix}\, \var{ms} \}\).
\end{lem}

@@

\begin{thm}[A linearized matrix has length the product of the shape]\label{Th:linearize_product}
    For any well-formed matrix \(m\), the length of the linearized contents is
    equal to the product of the computed shape of the matrix; that is,
    \(|\func{linearize'}\, (\func{contents}\, m)| = \prod
    \func{compute\_shape}\, (\func{contents}\, m)\).
\end{thm}

\begin{proof}
    We proceed by induction on the evidence for the inductive variant of
    the well-formed property (thanks to Theorem~\ref{Th:well_formed_agree}).

    The cases for \(\iname{WF\_Empty}\) and \(\iname{WF\_Scalar}\) are
    straightforward: computation via the definitions shows that both sides are
    euqal.

    In the inductive case for \(\iname{WF\_Cons}\), we combine a few facts about
    flat-map, length, and list concatenation with the inductive hypothesis to
    simplify the proof. The key piece of the proof is to show that the product
    of the computed shape of some \(\func{Matrix}\, \var{ms}\) is equal to the
    length of \(\var{ms}\) times the product of the computed shape of some \(m\)
    (both pieces of the evidence from the inductive judgement, and both
    with well-formed hypotheses). But this equality is straightforward by case
    analysis on \(\var{ms}\). The empty case is trivial. In the non-empty case,
    we know that the elements in the list and \(m\) have the same computed
    shaped because they are well-formed with respect to the same shape (see
    Lemma~\ref{Lem:wf_same_shape}). This is enough to finish.
\end{proof}

\begin{thm}[Linearize preserves the well-formed property]\label{Th:linearize_wf}
    For any well-formed matrix \(m\), the matrix \(\func{linearize}\, m\) is
    also well-formed.
\end{thm}

\begin{proof}
    We proceed by induction on the contents of \(m\), with the usual case
    analysis on the shape to eliminate contradictory cases.

    The base case is trivial by the definition of the well-formed property and
    the linearize function.

    In the inductive case, there are two major statements to prove. Firstly, we
    must prove that the length of the mapped \(\func{linearize'}\)d list is the
    product of the computed shape. Secondly, we must prove that all elements in
    the mapped \(\func{linearize'}\)d list are well-formed. See
    \S~\ref{S:linearize_coq} for the definition of \(\func{linearize}\) that
    leads to this mapped list.

    For the first statement, we use the fact that map does not change the length
    of the list together with Theorem~\ref{Th:linearize_product} to see that we
    only need to prove that a particular matrix is well-formed. For that proof,
    we combine Lemma~\ref{Lem:wfI_all_wf_t} with
    Theorem~\ref{Th:well_formed_agree} and the well-formed hypothesis.

    For the second statement, we only need to show that everything in the
    mapped \(\func{linearize'}\)d list is a \(\func{Scalar}\)---but we are
    mapping the \(\func{Scalar}\) constructor over the list, so this is of
    course true.
\end{proof}
